$\pi$ with leftovers: \\ a mechanisation in Agda

### 2021-01-06 paper/paper.tex
Get rid of REC

% Types
\newcommand{\Set}{\type{SET}}
\newcommand{\reduce}[1]{\, \type{\longrightarrow}_{#1} \,}
\newcommand{\types}[4]{#1 \, \type{;} \, #2 \, \type{\vdash} \, #3 \, \type{\triangleright} \, #4}
\newcommand{\contains}[6]{#1 \, \type{;} \, #2 \, \type{\ni}_{#3} \, #4 \, \type{;} \, #5 \, \type{\triangleright} \, #6}
\newcommand{\containsusage}[4]{#1 \, \type{\ni}_{#2} \, #3 \, \type{\triangleright} \, #4}
\newcommand{\Var}{\type{VAR}}
\newcommand{\Process}{\type{PROCESS}}
\newcommand{\Raw}{\type{RAW}}
\newcommand{\Name}{\type{NAME}}
\newcommand{\Names}{\type{NAMES}}
\newcommand{\Fresh}{\type{FRESH}}
\newcommand{\WellScoped}{\type{WELLSCOPED}}
\newcommand{\Unused}{\type{UNUSED}}
\newcommand{\PreCtx}{\type{PRECTX}}
\newcommand{\Ctx}{\type{CTX}}
\newcommand{\Type}{\type{TYPE}\,}
\newcommand{\Idx}{\type{IDX}\,}
\newcommand{\Idxs}{\type{IDXS}}
\newcommand{\Usage}{\type{USAGE}}
\newcommand{\N}{\type{\mathbb{N}}}
\newcommand{\Channel}{\type{CHANNEL}}
\newcommand{\Rec}{\type{REC}}
\newcommand{\Algebra}{\type{ALGEBRA}}
\newcommand{\eq}[1]{\, \type{\simeq}_{#1} \,}
\newcommand{\eqeq}{\, \type{\cong} \,}

@@

\begin{nidefinition}
We lift the relation \textsc{StructCong} $\eqeq{}$ and close it under equivalence and congruence in \textsc{Equals} $\eq{}$.
The $\constr{cong-}$* rules define structural congruence under a context $\mathcal{C}[\cdot]$ \cite{Sangio01}, respectively scope restriction, parallel composition, input and output.
The remaining three rules close structural congruence under reflexivity, symmetry and transitivity, respectively. 
The transitivity rule makes it tricky to prove predicates by induction on \textsc{Equals}: Agda cannot immediately see that such doubly recursive calls terminate.
In order to remedy this, we model the recursive call structure syntactically and index \textsc{Equals} by a type \textsc{Rec}.
\begin{mathpar}
  \datatype
  { }
  {\Rec : \Set}
  \; \textsc{Rec}

  \inferrule
  { }
  {\constr{zero} : \Rec}

  \inferrule
  {r : \Rec}
  {\constr{one} \; r : \Rec}

  \inferrule
  {r \; s : \Rec}
  {\constr{two} \; r \; s : \Rec}

  \datatype
  {P \, Q : \Process_n \\ r : \Rec}
  {P \eq{r} Q : \Set}
  \; \textsc{Equals}

  \inferrule
  {eq : P \eqeq Q}
  {\constr{struct} \; eq : P \eq{\constr{zero}} Q}

  \inferrule
  {eq : P \eq{r} P'}
  {\constr{cong-scope} \; eq : \new P \eq{\constr{one} \; r} \new P'}

  \inferrule
  {eq : P \eq{r} P'}
  {\constr{cong-comp} \; eq : \comp{P}{Q} \eq{\constr{one} \; r} \comp{P'}{Q}}

  \inferrule
  {eq : P \eq{r} P'}
  {\constr{cong-recv} \; eq : \recv{x}{}P \eq{\constr{one} \; r} \recv{x}{}P'}

  \inferrule
  {eq : P \eq{r} P'}
  {\constr{cong-send} \; eq : \send{x}{y}P \eq{\constr{one} \; r} \send{x}{y}P'}

  \inferrule
  { }
  {\constr{refl} : P \eq{\constr{zero}} P}

  \inferrule
  {eq : P \eq{r} Q}
  {\constr{sym} \; eq : Q \eq{\constr{one} \; r} P}

  \inferrule
  {eq_1 : P \eq{r} Q \\ \; eq_2 : Q \eq{s} R}
  {\constr{trans} \; eq_1 \; eq_2 : P \eq{\constr{two} \; r \; s} R}
\end{mathpar}
\end{nidefinition}

@@

We distinguish between channels that are created inside of the process ($\constr{internal}$), and channels that are created outside ($\constr{external}\ i$), where $i$ is the index of the channel variable.
In rule $\constr{comm}$, parallel processes reduce when they communicate over a common channel with index ${i}$.
As a result of that communication, the continuation of the input process $P$ has all the references to its most immediate variable substituted with references to $\suc j$, the variable sent by the output process $\send{i}{j}{Q}$.
After this substitution, $\subst{P}{\suc j}{\constr{0}}$ is \emph{lowered} --- all variable references are decreased by one (we apply \autoref{lm:substitution-unused} to obtain a proof $\Unused_{\constr{0}} \; (\subst{P}{\suc j}{\constr{0}})$).
Reduction is closed under parallel composition (rule $\constr{par}$), restriction (rule $\constr{res}$) and structural congruence (rule $\constr{struct}$) 
--- notably, not under input nor output, as doing so would not preserve the sequencing of actions \cite{Sangio01}.

Rule $\constr{res}$ uses $\func{dec}$ to decrement the index of channel $c$ as we wrap processes $P$ and $Q$ inside a binder.
It is defined as follows:
\begin{equation*}
  \begin{aligned}
    & \func{dec} : \Channel_(\suc n) && \to \Channel_n \\
    & \func{dec} \; \constr{internal}                 && \hookrightarrow \constr{internal} \\
    & \func{dec} \; (\constr{external} \; \constr{0}) && \hookrightarrow \constr{internal} \\
    & \func{dec} \; (\constr{external} \; (\suc n))   && \hookrightarrow \constr{external} \; n \\
  \end{aligned}
\end{equation*}
\todo{as inference rules, better}

@@

>>>>

% Types
\newcommand{\Set}{\type{SET}}
\newcommand{\reduce}[1]{\, \type{\longrightarrow}_{#1} \,}
\newcommand{\types}[4]{#1 \, \type{;} \, #2 \, \type{\vdash} \, #3 \, \type{\triangleright} \, #4}
\newcommand{\contains}[6]{#1 \, \type{;} \, #2 \, \type{\ni}_{#3} \, #4 \, \type{;} \, #5 \, \type{\triangleright} \, #6}
\newcommand{\containsusage}[4]{#1 \, \type{\ni}_{#2} \, #3 \, \type{\triangleright} \, #4}
\newcommand{\Var}{\type{VAR}}
\newcommand{\Process}{\type{PROCESS}}
\newcommand{\Raw}{\type{RAW}}
\newcommand{\Name}{\type{NAME}}
\newcommand{\Names}{\type{NAMES}}
\newcommand{\Fresh}{\type{FRESH}}
\newcommand{\WellScoped}{\type{WELLSCOPED}}
\newcommand{\Unused}{\type{UNUSED}}
\newcommand{\PreCtx}{\type{PRECTX}}
\newcommand{\Ctx}{\type{CTX}}
\newcommand{\Type}{\type{TYPE}\,}
\newcommand{\Idx}{\type{IDX}\,}
\newcommand{\Idxs}{\type{IDXS}}
\newcommand{\Usage}{\type{USAGE}}
\newcommand{\N}{\type{\mathbb{N}}}
\newcommand{\Channel}{\type{CHANNEL}}
\newcommand{\Rec}{\type{REC}}
\newcommand{\Algebra}{\type{ALGEBRA}}
\newcommand{\eq}{\, \type{\simeq} \,}
\newcommand{\eqeq}{\, \type{\cong} \,}

@@

\begin{nidefinition}
We lift the relation \textsc{StructCong} $\eqeq{}$ and close it under equivalence and congruence in \textsc{Equals} $\eq{}$.
The $\constr{cong-}$* rules define structural congruence under a context $\mathcal{C}[\cdot]$ \cite{Sangio01}, respectively scope restriction, parallel composition, input and output.
The remaining three rules close structural congruence under reflexivity, symmetry and transitivity, respectively. 
\begin{mathpar}
  \datatype
  {P \, Q : \Process_n \\ r : \Rec}
  {P \eq Q : \Set}
  \; \textsc{Equals}

  \inferrule
  {eq : P \eqeq Q}
  {\constr{struct} \; eq : P \eq Q}

  \inferrule
  {eq : P \eq P'}
  {\constr{cong-scope} \; eq : \new P \eq \new P'}

  \inferrule
  {eq : P \eq P'}
  {\constr{cong-comp} \; eq : \comp{P}{Q} \eq \comp{P'}{Q}}

  \inferrule
  {eq : P \eq P'}
  {\constr{cong-recv} \; eq : \recv{x}{}P \eq \recv{x}{}P'}

  \inferrule
  {eq : P \eq P'}
  {\constr{cong-send} \; eq : \send{x}{y}P \eq \send{x}{y}P'}

  \inferrule
  { }
  {\constr{refl} : P \eq P}

  \inferrule
  {eq : P \eq Q}
  {\constr{sym} \; eq : Q \eq P}

  \inferrule
  {eq_1 : P \eq Q \\ \; eq_2 : Q \eq R}
  {\constr{trans} \; eq_1 \; eq_2 : P \eq R}
\end{mathpar}
\end{nidefinition}

@@

We distinguish between channels that are created inside of the process ($\constr{internal}$), and channels that are created outside ($\constr{external}\ i$), where $i$ is the index of the channel variable.
In rule $\constr{comm}$, parallel processes reduce when they communicate over a common channel with index ${i}$.
As a result of that communication, the continuation of the input process $P$ has all the references to its most immediate variable substituted with references to $\suc j$, the variable sent by the output process $\send{i}{j}{Q}$.
After this substitution, $\subst{P}{\suc j}{\constr{0}}$ is \emph{lowered} --- all variable references are decreased by one (we apply \autoref{lm:substitution-unused} to obtain a proof $\Unused_{\constr{0}} \; (\subst{P}{\suc j}{\constr{0}})$).
Reduction is closed under parallel composition (rule $\constr{par}$), restriction (rule $\constr{res}$) and structural congruence (rule $\constr{struct}$) 
--- notably, not under input nor output, as doing so would not preserve the sequencing of actions \cite{Sangio01}.
\todo{why do we not apply structural congruence on both sides?}

@@

Rule $\constr{res}$ uses $\func{dec}$ to decrement the index of channel $c$ as we wrap processes $P$ and $Q$ inside a binder.
It is defined as follows:
\begin{equation*}
  \begin{aligned}
    & \func{dec} : \Channel_(\suc n) && \to \Channel_n \\
    & \func{dec} \; \constr{internal}                 && \hookrightarrow \constr{internal} \\
    & \func{dec} \; (\constr{external} \; \constr{0}) && \hookrightarrow \constr{internal} \\
    & \func{dec} \; (\constr{external} \; (\suc n))   && \hookrightarrow \constr{external} \; n \\
  \end{aligned}
\end{equation*}
\todo{introduce function definition syntax, fix alignment}
